### 2008-09-22 trunk/Papers/rmrs-08/rmrs.tex
version for Ann

We will now make the basic ideas from Section \ref{sec:motivation}
more precise.  We will first define the syntax of the \rmrs\ language;
this is a notational variant of earlier definitions in the
literature.  We will then define a model theory for our version of
\rmrs, and conclude this section by carrying over the notion of
\emph{solved forms} from CLLS \cite{egg:etal:2001}.

@@

We define the syntax of \rmrs\ in the style of CLLS
\cite{egg:etal:2001}.  We assume an infinite set of {\em node
  variables} $\nvar=\{X,Y,X_1,\ldots\}$, which we will use as labels,
anchors, and holes; the distinction between these uses will come from
their position in the formula.  Furthermore, we assume an infinite set
of \emph{base variables} $\ovar$, consisting of individual variables
$\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$.  Finally,
we assume a vocabulary of \emph{predicate symbols}
$\pred=\{P,Q,P_1,\ldots\}$, each of which has an \emph{arity} $\ar(P)
\geq 0$.

@@

A node variable $X$ is called a \emph{label} in the \rmrs\
$\varphi$ iff $\varphi$ contains an atom of the form $\ARG_S(X,i)$ or
an elementary predication of the form $\ep{Y}{X}{P}$; $X$ is called an
\emph{anchor} iff $\varphi$ contains an atom of the form
$\ep{X}{Y}{P}$ or $Y \qeq X$; and $X$ is called a \emph{hole} iff
$\varphi$ contains an atom of the form $\ARG_S(Y,X)$ or $X \qeq Y$.

@@

\rmrs\ as presented here combines similarities to earlier
presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
constraints \cite{egg:etal:2001}.  We $\ARG_{\{i\}}$ (with a singleton
set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
Furthermore, the elementary predication $\ep{l}{a}{P(v)}$ is a
notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.  In other
words, the syntax defined here differs in some details, but is more
general. The primary differences to dominance constraints are the
distinction of labels and anchors that will separate the labelling and
the parent-child relations, and the use of $\qeq$ instead of
dominance.  Redefining \rmrs\ in the style of dominance constraints
will simplify the formal work below.

\todo{we don't split stuff up into tuple parts, which earlier rmrs
  did}

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$, such as the one in Fig.~\ref{fig:1}.  So models of the
\rmrs\ are such trees, and we can define satisfaction and entailment
over these models in the usual way.  Satisfaction corresponds to an
\rmrs\ being a partial description of the (unique) fully specific
semantic representation that the model corresponds to, whereas
entailment corresponds to the relation between a more and a less
specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic; in this case, $\tau$ is simply the structure tree of
the predicate logic formula.  We assume that $\Sigma$ is a ranked
signature consisting of predicate logic's symbols: a unary constructor
$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; the
binary constructors $\forall x$ and $\exists x$ for all base-language
variables $x$; all the constants and base-language variables as
nullary constructors; and, for each predicate symbol $P$ of arity $n$,
a constructor $P$ of arity $n$.  Other base languages may require a
different signature $\Sigma$ and/or a different mapping between
formulas and trees.  We only assume for convenience that the signature
contains a binary constructor $\wedge$ to represent conjunction.  We
write $\Sigma_i$ for the set of all constructors in $\Sigma$ that have
arity $i$.  We will follow the typographical convention that
non-logical symbols in $\Sigma$ are written in
$\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
like ``$\sempred{\_cat\_n}$''.

@@

Notice that one \rmrs\ may be satisfied by multiple trees; we can take
the \rmrs\ to be a partial description of each of these trees.  In
particular, \rmrs s may represent semantic scope ambiguities, missing
information about semantic dependencies, and missing information about
lexical subcategorisation and lexical senses.  For $j=\{1,2\}$,
suppose that $\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let
$\nvar$ be the node variables from which $\varphi$ is built.  Then
$\varphi$ exhibits a semantic scope ambiguity if there are variables
$Y,Y'\in \nvar$ such that $\alpha_1(Y) \qeqdom \alpha_1(Y')$ and
$\alpha_2(Y')\qeqdom \alpha_2(Y)$.  It exhibits missing information
about semantic dependencies if there are base-language variables $
v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$.
It exhibits missing lexical subcategorisation information if there is
a variable $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
constructor of a different type from $\tau_2(\alpha_2(Y))$ \todo{what
  does this mean?}.  And it exhibits missing lexical sense information
if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$ are different
base-language constructors, but of the same type.

Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
This \rmrs\ is satisfied by the tree in Fig.~\ref{fig:1} (among
others) together with some particular $\alpha$, $g$, and $\sigma$.
The assignment $\alpha$ will map the variables $l_1$ and $a_1$ to the
root node; $l_{41}$ and $l_{42}$ to its second child (labeled with
``$\wedge$''); $a_{41}$ to the first child of \emph{that} node (i.e.\
the node 21, labelled with ``fat'') and $a_{42}$ to the node 22; and
so forth.  $g$ will map $x$ and $x_3$ to $x$, $e'$ to $e$, and so on.
And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
word) to the set of its word senses, e.g.\ \_cat\_n to a set
containing \_cat\_n\_1 and possibly others.  It is then easy to verify
that every single atom in the \rmrs\ is satisfied -- most
interestingly, the EPs $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``every fat black cat chased
some dog'', differs from the model in Fig.~\ref{fig:1} in the subtree
below the restrictor of the universal quantifier.  It is still a model
of (\ref{ex:cat-pos})---all the EPs are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula).  The \rmrs\
didn't specify that this material should be there; but it also didn't
specify that it can't be there, and indeed there is no way in which
the \rmrs\ can enforce that no extra material can be added.

@@

Intuitively, we expect that an \rmrs\ in solved form is fully
specified with respect to scope ambiguities and the predicate-argument
structure, all variable equalities and inequalities have been
resolved, and only lexical sense ambiguities remain.  This is made
precise below.

\todo{Note: We don't really talk about SPEC anywhere else.}

\begin{definition}\label{defn:solved-forms}
  An RMRS $\varphi$ is \emph{in solved form} iff:
  \begin{enumerate}
  \item every variable in $\varphi$ is either a hole, a label or an
    anchor (but not two of these);
  \item $\varphi$ doesn't contain equality, inequality, and SPEC
    ($\sqsubseteq$) atoms;
  \item if $\ARG_S(Y,i)$ is in $\varphi$, then $|S| = 1$;
  \item for any label $Y$ and index set $S$, there
    are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
  \item if $Y$ is an anchor in some EP $\ep{X}{Y}{P}$ and $k$ is the
    maximum number such that $\ARG_{\{k\}}(X,i)$ is in $\varphi$ for
    any $i$, then there is a constructor $p \in \sigma(P)$ whose arity
    is at least $k$;
  \item no label occurs on the right-hand side of two
    different $\qeqdom$ atoms.
  \end{enumerate}
\end{definition}

@@

\begin{prop} \label{prop:solved-forms-are-satisfiable}
  Every RMRS in solved form is satisfiable.
\end{prop}
\begin{proof}[Proof (sketch; see also
  \cite{Duchier00dominanceconstraints})] 
  For each EP, we choose to label the anchor with the constructor $p$
  of sufficiently high arity whose existence we assumed; we determine
  the edges between an anchor and its children from the uniquely
  determined $\ARG$ atoms; plugging labels into holes is
  straightforward because no label is qeq-dominated by more than one
  label; and then we can fill the spaces between the labels and
  anchors with conjunctions.
\end{proof}

@@

  An RMRS $\varphi'$ is \emph{a solved form of} the RMRS $\varphi$
  iff $\varphi'$ is in solved form and there is a substitution $s$
  that maps the  node and base variables of $\varphi$ to the node and
  base variables of $\varphi'$ such that
  \begin{enumerate}
  \item $\varphi'$ contains the EP $\ep{X'}{Y'}{P}$ iff there are
    variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
    s(X)$, and $Y' = s(Y)$;
  \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
    s(i)$, and $S' \subseteq S$;
  \item $D(\varphi') \supseteq s(D(\varphi))$.
  \end{enumerate}
\end{definition}

@@

\begin{prop}  \label{prop:finite-solved-forms}
  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
  renaming of variables.
\end{prop}
\begin{proof}
  Up to renaming of variables, there is only a finite number of
  substitutions on the node and base variables of $\varphi$.  Let $s$
  be such a substitution.  This fixes the set of EPs of any solved
  form of $\varphi$ that is based on $s$ uniquely.  There is only a
  finite set of choices for the subsets $S'$ in condition 2 of
  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
  the set of solved forms of $\varphi$ is finite.
\end{proof}

\subsection{Extensions}

>>>>

We will now make the basic ideas from Section \ref{sec:motivation}
more precise.  We will first define the syntax of the \rmrs\ language;
this is a notational variant of earlier definitions in the
literature.  We will then define a model theory for our version of
\rmrs, and conclude this section by carrying over the notion of
\emph{solved forms} from {\sc clls} \cite{egg:etal:2001}.

@@

We define the syntax of \rmrs\ in the style of {\sc clls}
\cite{egg:etal:2001}.  We assume an infinite set of {\em node
  variables} $\nvar=\{X,Y,X_1,\ldots\}$, which we will use as labels,
anchors, and holes; the distinction between these uses will come from
their position in the formula.  Furthermore, we assume an infinite set
of \emph{base variables} $\ovar$, consisting of individual variables
$\{x,x_1,y,\ldots\}$ and event variables $\{e_1,\ldots\}$.  Finally,
we assume a vocabulary of \emph{predicate symbols}
$\pred=\{P,Q,P_1,\ldots\}$, each of which has an \emph{arity} $\ar(P)
\geq 0$.

@@

A node variable $X$ is called a \emph{label} iff $\varphi$ contains an
atom of the form 
$\ep{X}{Y}{P}$ or $Y \qeq X$; $X$ is called an
 a \emph{anchor} in the \rmrs\
$\varphi$ iff $\varphi$ contains an at atom of the form
$\ep{Y}{X}{P}$ or $\ARG_S(X,i)$; and $X$ is
called a \emph{hole} iff 
$\varphi$ contains an atom of the form $\ARG_S(Y,X)$ or $X \qeq Y$.

@@

\rmrs\ as presented here combines similarities to earlier
presentations of \rmrs\ (e.g., \cite{copestake:2003}) and to dominance
constraints \cite{egg:etal:2001}.  We use $\ARG_{\{i\}}$ (with a singleton
set $S$) instead of $\ARG_i$, and $\ARG_\N$ instead of $\ARG_n$.
Furthermore, the elementary predication $\ep{l}{a}{P(v)}$ is a
notational variant of $\{\ep{l}{a}{P}, \ARG_{\{0\}}(l,v)\}$.  In other
words, the syntax defined here differs in some details, but is more
general. The primary differences to dominance constraints are the
distinction of labels and anchors that will separate the labelling and
the parent-child relations, and the use of $\qeq$ instead of
dominance.  Redefining \rmrs\ in the style of dominance constraints
will simplify the formal work below.

@@

The model theory formalises the relationship between an {\sc rmrs} and
the fully specific, alternative logical forms that it describes,
expressed in the base language.  We represent such a logical form as a
tree $\tau$, such as the ones in Fig.~\ref{fig:1}.  So models of the
\rmrs\ are such trees, and we can define satisfaction and entailment
over these models in the usual way.  Satisfaction corresponds to an
\rmrs\ being a partial description of the (unique) fully specific
semantic representation that the model corresponds to, whereas
entailment corresponds to the relation between a more and a less
specific \rmrs.

In this paper, we assume for simplicity that the base language is
predicate logic (e.g., see (\ref{ex:fat-cat-1}) and
(\ref{ex:fat-cat-2})); in this case, $\tau$ is simply the structure
tree of 
the predicate logic formula.  We assume that $\Sigma$ is a ranked
signature consisting of predicate logic's symbols: a unary constructor
$\neg$ and binary constructors $\wedge$, $\rightarrow$, etc.; the
binary constructors $\forall x$ and $\exists x$ for all base-language
variables $x$; all the constants and base-language variables are
nullary constructors; and, for each predicate symbol $P$ of arity $n$,
a constructor $P$ of arity $n$.  Other base languages may require a
different signature $\Sigma$ and/or a different mapping between
formulas and trees.  We only assume for convenience that the signature
contains a binary constructor $\wedge$ to represent conjunction.  We
write $\Sigma_i$ for the set of all constructors in $\Sigma$ that have
arity $i$.  We will follow the typographical convention that
non-logical symbols in $\Sigma$ are written in
$\mathsf{sans{-}serif}$, as opposed to the \rmrs\ predicate symbols
like ``$\sempred{\_cat\_n}$'' and ``$\sempred{\_cat\_n\_1}$''.

@@

Notice that one \rmrs\ may be satisfied by multiple trees; we can take
the \rmrs\ to be a partial description of each of these trees.  In
particular, \rmrs s may represent semantic scope ambiguities, missing
information about semantic dependencies, and missing information about
lexical subcategorisation and lexical senses.  For $j=\{1,2\}$,
suppose that $\tau_j,\alpha_j,g_j,\sigma\models \varphi$ and let
$\nvar$ and $\ovar$ be the node and base variables from which
$\varphi$ is built.  Then 
$\varphi$ exhibits a semantic scope ambiguity if there are variables
$Y,Y'\in \nvar$ such that $\alpha_1(Y) \qeqdom \alpha_1(Y')$ and
$\alpha_2(Y')\qeqdom \alpha_2(Y)$.  It exhibits missing information
about semantic dependencies if there are base-language variables $
v,v'\in \ovar$ such that $g_1(v)=g_1(v')$ and $g_2(v)\neq g_2(v')$.
It exhibits missing lexical subcategorisation information if there is
a variable $ Y \in \nvar$ such that $\tau_1(\alpha_1(Y))$ is a
constructor of a different type from $\tau_2(\alpha_2(Y))$  (i.e., the
constructors are of a different arity or and/or their arguments are).
And it exhibits missing lexical sense information
if $\tau_1(\alpha_1(Y))$ and $\tau_2(\alpha_2(Y))$ are different
base-language constructors, but of the same type.

Let's look at the \rmrs\ in (\ref{ex:cat-pos}) by way of example.
This \rmrs\ is satisfied by the trees in Fig.~\ref{fig:1} (among
others) together with some particular $\alpha$, $g$, and $\sigma$.
For instance, consider the left-hand side tree in Fig.~\ref{fig:1}.
The \rmrs\ (\ref{ex:cat-pos}) satisfies this tree with an assignment
function $\alpha$ that maps the variables $l_1$ and $a_1$ to the
root node, $l_{41}$ and $l_{42}$ to its second child (labeled with
``$\wedge$''), $a_{41}$ to the first child of \emph{that} node (i.e.\
the node 21, labelled with ``fat'') and $a_{42}$ to the node 22, and
so forth.  $g$ will map $x_1$ and $x_3$ to $x$, and $x_6$ and $x_7$ to
$y$, and so on.  
And $\sigma$ will map each \rmrs\ predicate symbol (which represents a
word) to the set of its word senses, e.g.\ \_cat\_n to a set
containing $\sem{\_cat\_n\_1}$ and possibly others.  It is then easy to verify
that every single atom in the \rmrs\ is satisfied---most
interestingly, the {\sc ep}s $\ep{l_{41}}{a_{41}}{\_fat\_a(e')}$ and
$\ep{l_{42}}{a_{42}}{\_cat\_n(x_3)}$ are satisfied because
$\alpha(l_{41}) \wedgedom \alpha(a_{41})$ and $\alpha(l_{42})
\wedgedom \alpha(a_{42})$.

@@

One feature which the above definition of \rmrs\ shares with dominance
constraints is that any satisfiable \rmrs\ has an infinite number of
models which only differ in the areas that the \rmrs\ didn't ``talk
about''.  An example is shown in Fig.~\ref{fig:fat-black-cat}.  This
tree, which represents the meaning of ``every fat black cat chased
some dog'', differs from the model in Fig.~\ref{fig:1} in the subtree
below the restrictor of the universal quantifier.  It is still a model
of (\ref{ex:cat-pos})---all the {\sc ep}s are satisfied.  However, there is
more ``semantic material'' (the ``black'' subformula).  The \rmrs\
didn't specify that this material should be there; but it also didn't
specify that it can't be there, and indeed there is no way in which
the \rmrs\ can enforce that no extra material can be added.

@@

Intuitively, we expect that an \rmrs\ in solved form is fully
specified with respect to the predicate-argument
structure, all variable equalities and inequalities have been
resolved, and only lexical sense ambiguities and (resolvable)
scope ambiguities remain.  This is made
precise below.

\begin{definition}\label{defn:solved-forms}
  An RMRS $\varphi$ is \emph{in solved form} iff:
  \begin{enumerate}
  \item every variable in $\varphi$ is either a hole, a label or an
    anchor (but not two of these);
  \item $\varphi$ doesn't contain equality, inequality, and SPEC
    ($\sqsubseteq$) atoms;
  \item if $\ARG_S(Y,i)$ is in $\varphi$, then $|S| = 1$;
  \item for any label $Y$ and index set $S$, there
    are no two atoms $\ARG_S(Y,i)$ and $\ARG_S(Y,i')$ in $\varphi$;
  \item if $Y$ is an anchor in some {\sc ep} $\ep{X}{Y}{P}$ and $k$ is the
    maximum number such that $\ARG_{\{k\}}(X,i)$ is in $\varphi$ for
    any $i$, then there is a constructor $p \in \sigma(P)$ whose arity
    is at least $k$;
  \item no label occurs on the right-hand side of two
    different $\qeqdom$ atoms.
  \end{enumerate}
\end{definition}

@@

\begin{prop} \label{prop:solved-forms-are-satisfiable}
  Every RMRS in solved form is satisfiable.
\end{prop}
\begin{proof}[Proof (sketch; see also
  \cite{Duchier00dominanceconstraints})] 
  For each {\sc ep}, we choose to label the anchor with the constructor $p$
  of sufficiently high arity whose existence we assumed; we determine
  the edges between an anchor and its children from the uniquely
  determined $\ARG$ atoms; plugging labels into holes is
  straightforward because no label is qeq-dominated by more than one
  label; and then we can fill the spaces between the labels and
  anchors with conjunctions.
\end{proof}

@@

  An RMRS $\varphi'$ is \emph{a solved form of} the RMRS $\varphi$
  iff $\varphi'$ is in solved form and there is a substitution $s$
  that maps the  node and base variables of $\varphi$ to the node and
  base variables of $\varphi'$ such that
  \begin{enumerate}
  \item $\varphi'$ contains the {\sc ep} $\ep{X'}{Y'}{P}$ iff there are
    variables $X,Y$ such that $\ep{X}{Y}{P}$ is in $\varphi$, $X' =
    s(X)$, and $Y' = s(Y)$;
  \item for every atom $\ARG_S(X,i)$ in $\varphi$, there is exactly
    one atom $\ARG_{S'}(X',i')$ in $\varphi$ with $X' = s(X)$, $i' =
    s(i)$, and $S' \subseteq S$;
  \item $D(\varphi') \supseteq s(D(\varphi))$.
  \end{enumerate}
\end{definition}

@@

\begin{prop}  \label{prop:finite-solved-forms}
  Every {\sc rmrs} $\varphi$ has only a finite number of solved forms, up to
  renaming of variables.
\end{prop}
\begin{proof}
  Up to renaming of variables, there is only a finite number of
  substitutions on the node and base variables of $\varphi$.  Let $s$
  be such a substitution.  This fixes the set of {\sc ep}s of any solved
  form of $\varphi$ that is based on $s$ uniquely.  There is only a
  finite set of choices for the subsets $S'$ in condition 2 of
  Def.~\ref{defn:solved-form-of}, and there is only a finite set of
  choices of new $\qeqdom$ atoms that satisfy condition 3.  Therefore,
  the set of solved forms of $\varphi$ is finite.
\end{proof}

\subsection{Extensions}
\label{sec:extensions}
